Visualizing demographic evolution using geographically inconsistent census data

### 2019-08-24 paper/intro.tex
almost done

\unsure{CHECK blue text below}
\revision{1.1}{In some cases, harmonisation might seem methodologically
adequate, if a study is focused on a small subregion, fixed in time, limiting
and modifying the data to fit that specific region of interest sounds not only
reasonable, but recommended, albeit the creation of artificial data points.
However, that is a simplistic view that ignores interactions beyond those
borders: each different portion of the area of interest may belong to larger
homogeneous processes beyond these arbitrarily defined borders. This effect,
along with the creation of artificial points, is caused by the geometry mismatch
between the data and the region of interest. Instead of forcing one set of
borders into another, our method sidesteps this issue by changing how the data
is represented and interpreted.

>>>>

\unsure{CHECK blue text below}
\revision{1.1}{In some cases, harmonisation might seem methodologically
adequate. If a study is focused on a small subregion that is fixed in time, limiting
and modifying the data to fit that specific region of interest sounds not only
reasonable, but recommended, even if it comes at the cost of producing artificial data points. However, this view is limited in that it ignores interactions beyond the fixed
borders: each different portion of the area of interest may belong to larger
homogeneous processes beyond these arbitrarily defined borders. This effect,
along with the creation of artificial points, is caused by the geometry mismatch
between the data and the region of interest. Instead of forcing one set of
borders into another, our method sidesteps this issue by changing how the data
is represented and interpreted. It therefore allows researchers to examine changes in both the composition and extent of local areas.

### 2019-08-24 paper/method.tex
almost done

@@

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, marked as \emph{who}
in Figure~\ref{fig:ui}. The stacked bar plots summarise the overall demographic
composition of these regions. In this example, the maps show the transition from
orange to green and purple in several regions over time. Each aspect is
represented by a stacked bar plot, where the width of each rectangle corresponds
to \revision{1.10}{the percentage of population in that variable over the
considered period. In this case, a little less than half of the population are
married, and the percentage that are Widowers or Divorced is roughly similar.
About half of the population work in Administrative jobs, a third never
completed high-school, approximately forty percent have gross family income
below 50,000USD per year. About sixty percent identify as white. Placing the
mouse over one of the bars will open a small panel with the temporal evolution
of the population percentage of that specific variable, and clicking on the
chevron on the right side expands the corresponding aspect, showing census tract
level statistics, with details of the temporal evolution of each variable and
also the corresponding IQRs for the whole city. Those values differ from the
total population percentage, while the overall population of this region is
about sixty percent White, the average percentage of White population over the
CTs is a little over forty percent.}

>>>>

The main functionalities of the interface allow users to examine four
inter-related questions central to understanding local area dynamics. 1)
\emph{What} do the clusters consist in?; 2) \emph{Where} are they located; 3)
\emph{How} do they change?; and 4) \emph{Who} occupy them? We discuss each in
turn.

@@

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, marked as \emph{who}
in Figure~\ref{fig:ui}. The stacked bar plots summarise the overall demographic
composition of these regions. In this example, the maps show the transition from
orange to green and purple in several regions over time. Each aspect is
represented by a stacked bar plot, where the width of each rectangle corresponds
to \revision{1.10}{the percentage of population in that variable over the
considered period. In this case, a little less than half of the population are
married, and the percentage that are Widowers or Divorced is roughly similar.
About half of the population work in Administrative jobs, a third never
completed high-school, approximately forty percent have gross family income
below 50,000USD per year. About sixty percent identify as white. Placing the
mouse over one of the bars will open a small panel with the temporal evolution
of the population percentage of that specific variable, and clicking on the
chevron on the right side expands the corresponding aspect, showing census tract
level statistics, with details of the temporal evolution of each variable and
also the corresponding IQRs for the whole city. Those values differ from the
total population percentage. While the overall population of this region is
about sixty percent White, the average percentage of White population over the
CTs is a little over forty percent.}

### 2019-08-24 paper/study.tex
almost done

The prototype interface allows access to 41 regions, 29
in the US and 12 in Canada. New York City was split into its boroughs to avoid
memory crashes on the client browser due to the high number of CTs.  We used
five aspects for the USA: Education level, Family income, Marital status,
Occupation, and Race; and seven for Canada: Age, Education level, Home language,
Household Income, Marital status, Occupation, Place of birth, and Religion.

While our method does not require geographic harmonisation, it requires matching
the variables over time. The supplementary material contains the details of
which census columns were used for each aspect. Income is slightly inaccurate,
even though we did correct for the official inflation. We grouped the original
ranges into three larger ranges, but they do not match precisely.

\unsure{CHECK}\revision{1.12.(1-5)}{These results are meant to demonstrate the
utility of the interface for understanding the evolutionary dynamics of urban
neighbourhoods. They also show the face validity of the results generated by our
novel network-based approach. However, the US data for 2010 is  not from the
decennial census, but from the ACS 2006-2010, compromising the temporal
stability of the data. Further, the regions selected do not correspond to any
pre-defined regions (metros, cities, census areas), but to somewhat arbitrary
regions defined around a location of interest, making direct comparisons harder.
Since  variable matching is a  manual process, we selected a minimal set of
aspects for each country, which are not necessarily similar to each other, or
similar to studies in the current literature. These factors would seriously
hinder any study using the existing methods, but our results were well-aligned
with several other studies, illustrating the robustness of our approach to
imperfect conditions. }

@@

The initial state of the prototype is illustrated in Figure~\ref{fig:ui}. The
first step is to identify the compositions of each cluster from the boxplots, so
orange is associated with majority of White population, green with majority
Black, and purple with higher proportion of four years of college or more (high
education level). The expanded version of the boxplots for the purple cluster
shows a higher income level and majority of occupations in administrative jobs,
therefore the purple cluster identifies gentrified regions.

The trajectories plot illustrates the process of gentrification, also
illustrated in Figure~\ref{fig:chiWorkflow}, progressively absorbing regions
from the orange cluster (White). This corroborates results from the literature
reporting that Black neighbourhoods are less likely to
gentrify~\citep{Hwang2014}. The transition matrix provides more details, showing
that around one percent of all regions that became purple at any time
transitioned from the green cluster (Black population), while nine percent were
previouly associated with the orange cluster (White). Next, we select the region
that is gentrified in 2010, by clicking on the corresponding rectangle in the
trajectories plot, updating the information on the maps and the details portion
of the interface.

The corresponding regions are highlighted in the maps, where the spatial pattern
is clear, corresponding exactly to previous findings in the literature based
upon harmonisation~\citep{Hwang2014}. Further, we can also identify the regions
that gentrified earlier on the small maps that depict the involved regions over
time. Since the most relevant aspect is Education, specifically "Four or more
years of college", we can expand the details of this aspect, as illustrated in
the rightmost portion of Figure~\ref{fig:chiWorkflow}, which is increasing for
the whole city (grey band), but faster and to a higher level in this region
(black band).

@@

The population with low percentage of University degrees is represented in
orange, mostly anglophone population in green, Asian immigrants in yellow, high
percentage of income in the highest bracket in purple, high percentage of Jewish
people in light green and brown, high percentage of Eastern Non-Christian
religion in pink, and high concentration of single people in dark grey. From the
trajectories plot, we can see that Toronto is more dynamic than Chicago, with
one cluster constantly shrinking. In the 1970s, the city was divided into four
clusters: low number of university degrees, Jewish population, majority
anglophones, and high income. Interestingly, the more recent clusters that
absorbed regions from the orange cluster have similar education profiles and are
differentiated by other aspects. In this sense, the city is growing diverse,
changing from a common low education profile to a higher level of education with
more diversity in religion (pink) and immigration (yellow).

Indeed, the growing Asian population is visible starting in the 1980s and
building thereafter, leading to the yellow and pink clusters. While both include
a high percentage of people born in Asia, the pink is more defined by religion,
with low percentage of university degrees, and contains the lowest percentage of
people in the highest income bracket for these clusters; the yellow is less
defined by religion, and has higher education and income, geographically
corresponding to the Markham region, know for its Chinese population. A similar
division also happens for the two Jewish clusters, where the light green cluster
has lower education and income levels than the brown cluster. The purple cluster
of high income is somewhat stable. Until 2011 the cluster included the Bridle
Path neighbourhood, known for its wealthy population. In 2011 it was classified
into the yellow cluster of Asian immigration, since about 40\% of the population
for this CT were then born in Asia. The income distribution did not change, with
85\% of the population with an income of 90k CAD or more.

The most significant indicator of Toronto's dynamism is the presence of grey
regions on the map. These represent regions associated to three or more clusters
over this five census period. Using the 'Add' mode for the trajectory selection,
we select their trajectories, and a subset of the details is illustrated in
Figure~\ref{fig:toVol}. These regions account for about 5\% of Toronto's
population. The whole region was classified into the orange cluster in 1971 (low
level of university degrees). By 1991, most of the region was classified into
the green cluster, representing anglophone population, mostly Canadian born,
with a higher level of education. As the corresponding plot indicates, this
trend in increasing education is city-wide, but this region has people with
better education than most.

@@

In 2001, the purple cluster of high income annexes neighbouring parts of the
volatile region, and the Asian born population increases sharply, as illustrated
by the appearance of the yellow cluster.  This cluster indicates well educated,
higher income, and about 30\%-50\% Asian born population. By 2011, the yellow
cluster increased considerably, annexing parts of the high income purple
cluster, including the neighbouring Bridle Path area.

>>>>

The prototype interface allows access to 41 regions, 29
in the US and 12 in Canada. New York City was split into its boroughs to avoid
memory crashes on the client browser due to the high number of CTs.  We used
five aspects for the USA: Education level, Family income, Marital status,
Occupation, and Race; and seven for Canada: Age, Education level, Home language,
Household Income, Marital status, Occupation, Place of birth, and Religion. The supplementary material contains the details of
which census columns were used for each aspect.

\unsure{CHECK}\revision{1.12.(1-5)}{These results are meant to demonstrate the
utility of the interface for understanding the evolutionary dynamics of urban
neighbourhoods. They also show the face validity of the results generated by our
novel network-based approach. While our method does not require geographic harmonisation, it does however require matching
the variables over time. This is (currently) a painstaking, manual process, and the current version of the interface remains somewhat limited. For example, we selected a minimal set of
aspects for each country, which are not necessarily similar to each other, or
similar to studies in the current literature. Income moreover is slightly inaccurate,
even though we did correct for official inflation. We grouped the original
ranges into three larger ranges, but they do not match precisely. In addition, the US data labelled "2010" is  not from the decennial census, but from the ACS 2006-2010, somewhat compromising the temporal
stability of the data. Further, the regions selected do not correspond to any
pre-defined regions (metros, cities, census areas), but to somewhat arbitrary
regions defined around a location of interest, making direct comparisons harder.
Taken together, these factors would no doubt require caveats for 
a substantive study using the current prototype version of the interface. Nevertheless, in the present context we find that the interface results are well-aligned
with several other studies, illustrating the robustness of our approach to
imperfect conditions. }

@@

The initial state of the prototype is illustrated in Figure~\ref{fig:ui}. The
first step is to identify \textit{what} the clusters consist in. To do so, we
examine the compositions of each cluster from the boxplots. Orange is associated
with majority of White population, green with majority Black, and purple with
higher proportion of four years of college or more (high education level). The
expanded version of the boxplots for the purple cluster shows a higher income
level and majority of occupations in administrative jobs. The purple cluster
therefore could be described as gentrified regions.

The trajectories plot shows \textit{how} clusters have changed over time.  For
example, it illustrates the process of gentrification, also illustrated in
Figure~\ref{fig:chiWorkflow}, with the purple cluster progressively absorbing
regions from the orange cluster (White). This corroborates results from the
literature reporting that Black neighbourhoods are less likely to
gentrify~\citep{Hwang2014}. The transition matrix provides more details, showing
that around one percent of all regions that became purple at any time
transitioned from the green cluster (Black population), while nine percent were
previouly associated with the orange cluster (White). Next, we select the region
that is gentrified in 2010, by clicking on the corresponding rectangle in the
trajectories plot, updating the information on the maps and the details portion
of the interface.

The maps show \textit{where} these processes are occurring, by highlighting the
corresponding regions. The spatial pattern is clear, corresponding exactly to
previous findings in the literature based upon harmonisation~\citep{Hwang2014}.
Further, we can also identify the regions that gentrified earlier on the small
maps that depict the involved regions over time. Finally, we can examine
\textit{who} is driving these changes. Since the most relevant aspect is
Education, specifically "Four or more years of college", we can expand the
details of this aspect, as illustrated in the rightmost portion of
Figure~\ref{fig:chiWorkflow}, which is increasing for the whole city (grey
band), but faster and to a higher level in this region (black band).

\unsure{CHECK - maybe add what/how/where/who to figure 5?}

@@

We again are guided by the four questions embedded in the interface: what, how,
where, and who. The boxplots address what clusters defined Toronto over this
period. The population with low percentage of University degrees is represented
in orange, mostly anglophone population in green, Asian immigrants in yellow,
high percentage of income in the highest bracket in purple, high percentage of
Jewish people in light green and brown, high percentage of Eastern Non-Christian
religion in pink, and high concentration of single people in dark grey.

From the trajectories plot, we can see how the city changed. It shows that
Toronto is more dynamic than Chicago, with one cluster constantly shrinking. In
the 1970s, the city was divided into four clusters: low number of university
degrees, Jewish population, majority anglophones, and high income.
Interestingly, the more recent clusters that absorbed regions from the orange
cluster have similar education profiles and are differentiated by other aspects.
In this sense, the city is growing diverse, changing from a common low education
profile to a higher level of education with more diversity in religion (pink)
and immigration (yellow). Indeed, the growing Asian population is visible
starting in the 1980s and building thereafter, leading to the yellow and pink
clusters. While both include a high percentage of people born in Asia, the pink
is more defined by religion, with low percentage of university degrees, and
contains the lowest percentage of people in the highest income bracket for these
clusters; the yellow is less defined by religion, and has higher education and
income, geographically corresponding to the Markham region, know for its Chinese
population. A similar division also happens for the two Jewish clusters, where
the light green cluster has lower education and income levels than the brown
cluster. The purple cluster of high income is somewhat stable. Until 2011 the
cluster included the Bridle Path neighbourhood, known for its wealthy
population. In 2011 it was classified into the yellow cluster of Asian
immigration, since about 40\% of the population for this CT were then born in
Asia. The income distribution did not change, with 85\% of the population with
an income of 90k CAD or more.

Examining where and by whom these changes occurred reveals additional
information about the city's underlying dynamics. The most significant indicator
of Toronto's dynamism is the presence of grey regions on the map. These
represent regions associated to three or more clusters over this five census
period. Using the 'Add' mode for the trajectory selection, we select their
trajectories, and a subset of the details is illustrated in
Figure~\ref{fig:toVol}. These regions account for about 5\% of Toronto's
population. The whole region was classified into the orange cluster in 1971 (low
level of university degrees). By 1991, most of the region was classified into
the green cluster, representing anglophone population, mostly Canadian born,
with a higher level of education. As the corresponding plot indicates, this
trend in increasing education is city-wide, but this region has people with
better education than most. In 2001, the purple cluster of high income annexes
neighbouring parts of the volatile region, and the Asian born population
increases sharply, as illustrated by the appearance of the yellow cluster.  This
cluster indicates well educated, higher income, and about 30\%-50\% Asian born
population. By 2011, the yellow cluster increased considerably, annexing parts
of the high income purple cluster, including the neighbouring Bridle Path area.

@@
